---
title: "Fitting GP to lightcurve 428_...in Stan"
format: html
---

Notebook outlining the fitting of GP to thunderKAT lightcurve ID$ 428_...

```{r setup}
library(conflicted)
library(ggplot2)
library(readr)
library(here)
library(cmdstanr)
library(posterior)
library(bayesplot)

color_scheme_set("brightblue")
register_knitr_engine(override = FALSE)
```

```{r fig.width=9}
csv_fname <- "428_ra236.530_dec-46.922_4U1543TraPDB_andersson.csv"
path_to_csv <- here("data_raw", csv_fname)
data_df <- read_csv(path_to_csv, show_col_types = FALSE)

data_df |> 
  ggplot() +
  aes(x = mjd, y = f_peak, ymax = f_peak + f_peak_err, ymin = f_peak - f_peak_err) +
  geom_point(colour = "blue") +
  geom_linerange() +
  labs(title = csv_fname, x = "MJD", y = "Flux Density") +
  theme_classic()
```

## Base Model

- Zero mean function
- Homoskedastic noise
- Not using data on error in observations
- Wide prior on observational errors

$$y \sim \mathcal{N}(f(x), \sigma_\textrm{noise}^2)$$

$$f \sim \mathcal{GP}(\boldsymbol{0}, k(x,  x'))$$


$$k(x,x') = \eta^2 \exp\left\{ -\frac{1}{2}\frac{(x - x')^2}{\ell^2}\right\}$$

$$\ell \sim \mathrm{InvGamma}(5,5)$$

$$\eta \sim \mathcal{N}^+(0,1)$$

$$\sigma_\textrm{noise} \sim \mathcal{N}^+(0,1)$$

```{cmdstan, cache=TRUE, output.var = "base_model"}
data {
  int<lower=1> N;
  array[N] real x;
  vector[N] y;
}
transformed data {
  vector[N] mu = rep_vector(0, N);
}
parameters {
  real<lower=0> ell;
  real<lower=0> eta;
  real<lower=0> sigma;
}
model {
  matrix[N, N] K = gp_exp_quad_cov(x, eta, ell);
  matrix[N, N] L = cholesky_decompose(add_diag(K, sigma^2));

  ell ~ inv_gamma(5, 5);
  eta ~ std_normal();
  sigma ~ std_normal();

  y ~ multi_normal_cholesky(mu, L);
}
```

```{r message=FALSE}
data_list <- list(N = length(data_df$mjd), 
                  x = data_df$mjd,
                  y = data_df$f_peak)

base_fit <- base_model$sample(
  data = data_list,
  seed = 1,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
```

```{r}
base_fit$cmdstan_diagnose()
```

```{r}
base_fit$print(variables = c("eta", "ell", "sigma"), digits = 6)
```

```{r}
base_draws_arr <- base_fit$draws(format = "draws_array")
mcmc_trace(base_draws_arr, pars = c("eta", "ell", "sigma"), facet_args = list(nrow = 2))
```

```{r}
mcmc_dens(base_draws_arr, pars = c("eta", "ell", "sigma"), facet_args = list(nrow = 2))
```

```{r}
mcmc_pairs(base_draws_arr, pars = c("eta", "ell", "sigma"), off_diag_fun = "hex")
```

## Observational Errors Model

- Zero mean function
- Include data on error in observations of $y_i$
- Heteroskedastic (Gaussian) noise

$$y_i \sim \mathcal{N}(f(x_i), \sigma_i^2)$$

$$f \sim \mathcal{GP}(\boldsymbol{0}, k(x,  x'))$$

$$k(x,x') = \eta^2 \exp\left\{ -\frac{1}{2}\frac{(x - x')^2}{\ell^2}\right\}$$

$$\ell \sim \mathrm{InvGamma}(5,5)$$

$$\eta \sim \mathcal{N}^+(0,1)$$

$$\sigma_i \sim \mathcal{N}^+(\textrm{stderr}(y_i), \mathrm{Var}(\textrm{stderr}(\boldsymbol{y})))$$

```{cmdstan, cache=TRUE, output.var = "err_model"}
functions {
  vector gp_pred_rng(array[] real x2,
                     vector y1,
                     array[] real x1,
                     real eta,
                     real ell,
                     vector sigma,
                     real jitter) {
    int N1 = rows(y1);
    int N2 = size(x2);
    vector[N2] f2;
    {
      matrix[N1, N1] L_K;
      vector[N1] K_div_y1;
      matrix[N1, N2] k_x1_x2;
      matrix[N1, N2] v_pred;
      vector[N2] f2_mu;
      matrix[N2, N2] cov_f2;
      matrix[N1, N1] K;
      K = gp_exp_quad_cov(x1, eta, ell);
      for (n in 1:N1)
        K[n, n] = K[n,n] + square(sigma[n]);
      L_K = cholesky_decompose(K);
      K_div_y1 = mdivide_left_tri_low(L_K, y1);
      K_div_y1 = mdivide_right_tri_low(K_div_y1', L_K)';
      k_x1_x2 = gp_exp_quad_cov(x1, x2, eta, ell);
      f2_mu = (k_x1_x2' * K_div_y1);
      v_pred = mdivide_left_tri_low(L_K, k_x1_x2);
      cov_f2 = gp_exp_quad_cov(x2, eta, ell) - v_pred' * v_pred;

      f2 = multi_normal_rng(f2_mu, add_diag(cov_f2, rep_vector(jitter, N2)));
    }
    return f2;
  }
}
data {
  int<lower=1> N;
  array[N] real x;
  vector[N] y;
  vector[N] y_stderr;
  int<lower=1> N_star;
  array[N_star] real x_star;
}
transformed data {
  vector[N] mu = rep_vector(0, N);
}
parameters {
  real<lower=0> ell;
  real<lower=0> eta;
  vector<lower=0>[N] sigma;
}
model {
  matrix[N, N] K = gp_exp_quad_cov(x, eta, ell);
  matrix[N, N] L = cholesky_decompose(add_diag(K, sigma^2));

  ell ~ inv_gamma(5, 5);
  eta ~ std_normal();
  sigma ~ normal(y_stderr, sd(y_stderr));

  y ~ multi_normal_cholesky(mu, L);
}
generated quantities {
  vector[N_star] f_star = gp_pred_rng(x_star, y, x, eta, ell, sigma, 1e-9);
}
```

```{r message=FALSE}
x_star <- seq(from = min(data_df$mjd), to = max(data_df$mjd), length.out = 200)

data_list <- list(N = length(data_df$mjd), 
                  x = data_df$mjd,
                  y = data_df$f_peak,
                  y_stderr = data_df$f_peak_err,
                  x_star = x_star,
                  N_star = length(x_star)
                  )

err_fit <- err_model$sample(
  data = data_list,
  seed = 1,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
```

```{r}
err_fit$cmdstan_diagnose()
```

```{r}
err_fit$print(variables = c("eta", "ell", "sigma"), digits = 6, max_rows = 23)
```

```{r}
err_draws_arr <- err_fit$draws(format = "draws_array")
mcmc_trace(err_draws_arr, pars = c("eta", "ell"), facet_args = list(nrow = 2))
```

```{r}
mcmc_dens(err_draws_arr, pars = c("eta", "ell"), facet_args = list(nrow = 2))
```

```{r}
mcmc_pairs(err_draws_arr, pars = c("eta", "ell"), off_diag_fun = "hex")
```

```{r fig.height=12}
mcmc_dens(err_draws_arr, regex_pars = "sigma", facet_args = list(nrow = 7))
```

```{r fig.width=9}
post_pred_draws <- as_draws_rvars(err_fit$draws("f_star"))

ggplot() +
  aes(x = x_star) +
  geom_ribbon(aes(ymin = quantile(post_pred_draws$f_star, probs = 0.16)[1,],
                  ymax = quantile(post_pred_draws$f_star, probs = 0.84)[1,]),
              fill = "blue", alpha = 0.3) +
  geom_ribbon(aes(ymin = quantile(post_pred_draws$f_star, probs = 0.05)[1,],
                  ymax = quantile(post_pred_draws$f_star, probs = 0.95)[1,]),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = median(post_pred_draws$f_star)), size = 1, colour = "black") +
  geom_point(aes(x = data_df$mjd, y = data_df$f_peak), size = 2, colour = "red") +
  labs(x = "MJD", y = "Flux Density") +
  theme_classic()
```


## Non-zero Mean function

- Constant mean function
- Include data on error in observations of $y_i$
- Heteroskedastic (Gaussian) noise

$$y_i \sim \mathcal{N}(f(x_i), \sigma_i^2)$$

$$f \sim \mathcal{GP}(C, k(x,  x'))$$

$$C \sim \mathcal{U}[0,1]$$

$$k(x,x') = \eta^2 \exp\left\{ -\frac{1}{2}\frac{(x - x')^2}{\ell^2}\right\}$$

$$\ell \sim \mathrm{InvGamma}(5,5)$$

$$\eta \sim \mathcal{N}^+(0,1)$$

$$\sigma_i \sim \mathcal{N}^+(\textrm{stderr}(y_i), \mathrm{Var}(\textrm{stderr}(\boldsymbol{y})))$$

